---
title: "Customer Behaviour Segmentation"
output:
  pdf_document: default
  html_notebook: default
  word_document: default
  html_document:
    df_print: paged
---

This study explores customer behavior segmentation using three distinct clustering models; K-Means, DBSCAN and Hierarchical Clustering in R-studio applied to the Mall Customer Dataset. The primary objective is to understand and categorize customer behavior based on their spending patterns and demographic attributes and suggest the best model for customer segmentation and recommendations. .

The dataset, containing information such as customer age, annual income, and spending score, undergoes preprocessing steps like scaling or normalization. Subsequently, DBSCAN, Hierarchical Clustering, and K-Means algorithms are applied to cluster customers into distinct segments.

Evaluation metrics and visualizations, including scatter plots, pair plots, and cluster comparison charts, are utilized to assess the quality and effectiveness of the generated clusters. The study compares the performance of these models in identifying coherent customer segments, highlighting their strengths and weaknesses in handling varying shapes and sizes of clusters, noise, and outliers within the dataset.

This investigation aims to provide insights into the diverse approaches of customer segmentation using different clustering methodologies, offering guidance on the most suitable model for extracting meaningful and actionable customer behavior patterns from the Mall Customer Dataset. 

```{r}
# Load dataset 
customer_data <- read.csv("C:/Users/LENOVO/Desktop/ThinkPacificProject/Mall_Customers.csv")
```

```{r}
#Rename some column names
customer_data <- rename(customer_data, Annual_Income=Annual.Income..k..,
                   Spending_Score=Spending.Score..1.100., Gender=Genre)
```

```{r}
#View customer
View(customer_data)
```

```{r}
# Convert heart_data to a data frame
customer_df <- as.data.frame(customer_data)
```

```{r}
# make a copy of the df
customer_df_km <- customer_df
```

```{r}
# view the head
head(customer_df)
```

```{r}
#dimensions of the data
dim(customer_df)

```

```{r}
#data structure
str(customer_df)
```


```{r}
# list types for each attribute
sapply(customer_df, class)

```
```{r}
#data summary
summary(customer_df)

```


```{r}
#checking for missing values
sum(is.na(customer_df))
```

```{r}
# Check missing values for each of the columns
customer_df %>%
  purrr::map_df(~ sum(is.na(.)))

```
```{r}
# View summary statistics with numeric hist
skimmed_data <- skim(customer_df)
View(skimmed_data)
```

```{r}
# Count the occurrences of unique values in the "target" column
table(customer_df$Gender)

```
## Exploratory Data Analysis(EDA)

```{r}
# Creating a barplot to assess gender distribution of my sample of customers.
ggplot(customer_df,aes(x= Gender)) +
  geom_bar(stat="count",width=0.5,fill="steelblue") +
  theme_minimal()+
  labs(title="Barplot to display Gender Comparison", xlab="Gender")

```

```{r}
# Age distribution of the customers
ggplot(customer_df, aes(x = Age)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Customer Ages",
       x = "Age",
       y = "Count") +
  theme_minimal()
```

```{r}
# Annual income distribution of the customers
ggplot(customer_df, aes(x = Annual_Income)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Annual Income",
       x = "Annual Income",
       y = "Count") +
  theme_minimal()
```

```{r}
# Annual income distribution of the customers
ggplot(customer_df, aes(x = Spending_Score)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Spending Score",
       x = "Spending Score",
       y = "Count") +
  theme_minimal()

```

```{r}
# Calculate the correlation matrix
numeric_customer_df <- customer_df[, sapply(customer_df, is.numeric)]
corr_matrix <- cor(numeric_customer_df)

# Heatmap 
ggplot(data = melt(corr_matrix), aes(Var1, Var2, fill = value)) +
  geom_tile() +
  geom_text(aes(label = round(value, 2)), vjust = 1) +  # Add labels with 2 decimal places
  scale_fill_gradient(low = "steelblue", high = "gray") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Correlation Heatmap")

```
```{r}
# Create a histogram with ggplot
ggplot(customer_df, aes(x = Age, fill = Gender)) +
  geom_histogram(binwidth = 1, color = 'black', alpha = 0.7) +
  labs(title = 'Age Distribution of Customers by Gender',
       x = 'Age',
       y = 'Count') +
  theme_minimal() +
  scale_fill_manual(values = c("gray", 'steelblue'))

```

```{r}

# Histogram to show dispersion of mall customers based on age groups
ggplot(customer_df, aes(x = Age)) +
  geom_vline(aes(xintercept = mean(Age)), color = "black",  #adding an intercept to indicate mean age
             linetype = "dashed", size = 1.0) +
  geom_histogram(binwidth = 5, aes(y = ..density..), 
                 color = "black", fill = "steelblue") +
  geom_density(alpha = 0.4, fill = "gray") +  #adding density plot
  labs(title = "Histogram to Show Density of Age Groups")

```


```{r}
# Creating density plot to show customer's annual income
ggplot(customer_df, aes(x = Annual_Income)) +
  geom_density(alpha=0.4, fill="steelblue") +
  scale_x_continuous(breaks = seq(0, 200, by = 10)) +
  labs(title="Density Plot for Annual Income")

```

```{r}
# Assuming 'data' is your data frame
plot(customer_df$CustomerID, customer_df$Annual_Income, 
     type = "p", 
     col = "steelblue", 
     main = "Annual Income by CustomerID", 
     xlab = "CustomerID", 
     ylab = "Annual Income"
)
```

```{r}

# Assuming 'customer_df' is your data frame
ggplot(customer_df, aes(x = Age, y = Annual_Income, color = Gender)) +
  geom_point(size = 3) +
  labs(title = "Scatter Plot of Age vs Annual Income by Gender",
       x = "Age",
       y = "Annual Income") +
  scale_color_manual(values = c("orange","steelblue")) +
  theme_minimal()

```

```{r}
# Assuming 'customer_df' is your data frame
ggplot(customer_df, aes(x = Age, y = Spending_Score, color = Gender)) +
  geom_point(size = 3) +
  labs(title = "Scatter Plot of Age vs Spending Score by Gender",
       x = "Age",
       y = "Spending Score") +
  scale_color_manual(values = c("orange","steelblue")) +
  theme_minimal()

```

```{r}
# Assuming 'customer_df' is your data frame
ggplot(customer_df, aes(y = Spending_Score, x = Annual_Income, color = Gender)) +
  geom_point(size = 3) +
  labs(title = "Scatter Plot of Annual Income vs Spending Score by Gender",
       y = "Spending Score",
       x = "Annual Income") +
  scale_color_manual(values = c("orange","steelblue")) +
  theme_minimal()

```

```{r}
# Assuming 'customer_df' is your data frame
ggplot(customer_df, aes(x = Spending_Score, y = Gender, fill = Gender)) +
  geom_boxplot() +
  labs(title = "Boxplot showing customers' Spending Score by Gender") +
  scale_fill_manual(values = c("orange","steelblue")) +
  theme_minimal()

```
## Training the Models

## Model Implementation Using K-Means Clustering

```{r}
# Setting seed to 125 for reproducibility
set.seed(1)

```

```{r}
# Plotting the variations of clusters 
opt <- Optimal_Clusters_KMeans(customer_df[, 3:5], max_clusters = 10, plot_clusters = T)

```

```{r}
# Getting the optimal clusters 
opt <- Optimal_Clusters_KMeans(customer_df[, 3:5], max_clusters = 10, plot_clusters = T, criterion = 'silhouette')

```


```{r}
#Creating the customer clusters with KMeans
km6<-kmeans(customer_df[,3:5], 6, iter.max = 100, nstart=50,
           algorithm = "Lloyd")
#Print the result
km6
```

```{r}
# Define the clustering variable
cluster_variable <- km6$cluster

# Showing the six KMeans clusters
clusplot(
  customer_df,                    # Data frame
  cluster_variable,               # Clustering variable
  color = TRUE,                   # Use color for clusters
  shade = TRUE,                   # Use shading
  labels = 0,                     # Do not show point labels
  lines = 0,                      # Do not show lines
  main = "Cluster Plot of Customer Behaviour",  # Main title
  sub = paste("Number of Clusters:", max(cluster_variable)),  # Subtitle with the number of clusters
  col.p = "steelblue",            # Color for points
  col.clus = c("darkred", "green", "blue", "orange"),  # Colors for cluster ellipses
  cex = 0.8                       # Size of labels and points
)

```


```{r}

#Perform Principal Component Analysis
pcclust<-prcomp(customer_df[, 3:5], scale=FALSE)

#Checking the summary of the PCA model
summary(pcclust)

```

```{r}
# Applying the PCA model on the data
pcclust$rotation[, 1:2]

```

```{r}
# A plot of the customers segments

ggplot(customer_df, aes(x = Annual_Income , y = Spending_Score)) + 
  geom_point(stat = "identity", aes(color = as.factor(km6$cluster))) +
  scale_color_discrete(name = " ", 
                       breaks=c("1", "2", "3", "4", "5","6"),
                       labels=c("Cluster 1", "Cluster 2", "Cluster 3", 
                                "Cluster 4", "Cluster 5","Cluster 6")) +
  ggtitle("Segments of Mall Customers", 
          subtitle = "Using K-means Clustering")

```

```{r}
# A plot of the customers segments
ggplot(customer_df, aes(x = Annual_Income , y = Spending_Score)) + 
  geom_point(stat = "identity", aes(color = as.factor(km6$cluster))) +
  scale_color_discrete(name = " ", 
                       breaks=c( "6", "1", "4", "3", "2","5"),
                       labels=c("Low Income, High Spending", "Medium Income, Medium Spending", "Low Income, Low Spending",
                                "Medium Income, Medium Spending","High Income, Low Spending", "High Income, High Spending")) +
  labs(x="Annual Income", y="Spending Score") +
  ggtitle("Segments of Mall X Customers", 
          subtitle = "Using K-means Clustering")

```

## Model Implementation Using DBSCAN clustering


```{r}
# Extracting the columns to be standardized
columns_to_standardize <- c("Age", "Annual_Income", "Spending_Score")
customer_df_scaled <- customer_df

# Standardizing the selected columns in the dataframe
customer_df_scaled[, columns_to_standardize] <- scale(customer_df[, columns_to_standardize])

```

```{r}
# Run DBSCAN
dbscan_result <- dbscan(customer_df_scaled[, c("Age", "Annual_Income", "Spending_Score")], eps = 0.5, minPts = 5)

```

```{r}
# Access cluster assignments and noise points
clusters <- dbscan_result$cluster
noise_points <- dbscan_result$cluster == 0  # Noise points will have cluster ID 0

```

```{r}
# Convert to a data frame for ggplot
customer_df_scaled$Cluster <- as.factor(clusters)
```

```{r}
# Visualize clusters and noise points using ggplot
ggplot(customer_df_scaled, aes(x = Annual_Income, y = Spending_Score, color = Cluster)) +
  geom_point() +
  geom_point(data = subset(customer_df_scaled, noise_points), color = "black", size = 3) +
  labs(x = "Annual Income", y = "Spending Score", title = "DBSCAN Clustering") +
  scale_color_discrete(name = " ", 
                       breaks=c("1", "3", "2", "4", "6","5"),
                       labels=c("Low Income, High Spending", "Medium Income, Medium Spending", "Low Income, Low Spending",
                                "Medium Income, Medium Spending","High Income, Low Spending", "High Income, High Spending")) +
  theme_minimal()


```


## Model Implementation Using Hierarchical clustering


```{r}
# Comparing the distances between the data points using ‘hclust’ function
dist_customers <- dist(customer_df[,3:5])
hc_customers <- hclust(dist_customers, method = 'complete')

```

```{r}
# Choosing the optimal cluster
for(i in 2:7) { 
  nam <- paste("clust", i, sep = "")
  assign(nam, cutree(hc_customers, k = i))
}

par(mfrow = c(3, 2))

plot(silhouette(clust2, dist_customers), col = "blue")
plot(silhouette(clust3, dist_customers), col = 'blue')
plot(silhouette(clust4, dist_customers), col = "blue")
plot(silhouette(clust5, dist_customers), col = "blue")
plot(silhouette(clust6, dist_customers), col = "blue")
plot(silhouette(clust7, dist_customers), col = "blue")


```

```{r}
# plot the dendrogram.
clust_customers <- cutree(hc_customers, k = 5)
dend_customers <- as.dendrogram(hc_customers)
dend_colored <- color_branches(dend_customers, k = 5)
par(mfrow = c(1, 1))
plot(dend_colored)

```

```{r}
# Extract the clusters
km <- kmeans(customer_df[,3:4], 5)
customer_df_km$ClusterNumber <- km$cluster
segment_customers <- mutate(customer_df_km, cluster = clust_customers)
segment_customers = subset(segment_customers, select = -c(ClusterNumber))
```

```{r}
#Create a plot of the customers segments
ggplot(segment_customers, aes(x = Annual_Income, y = Spending_Score, color = factor(cluster))) +
  geom_point() +
  scale_color_discrete(name = " ", 
                       breaks=c("1", "2", "3", "4", "5"),
                       labels=c("Low Income, Low Spending","Low Income, High Spending", "Medium Income, Medium Spending","High Income, High Spending",
                                "High Income, Low Spending")) +
  labs(x="Annual Income", y="Spending Score") +
  ggtitle("Segments of Mall X Customers", 
          subtitle = "Using Hierarchical Clustering")

```

```{r}
# Getting the characteristcs each of the groups
segment_customers %>% group_by(cluster, Gender) %>%
  summarise_all(list(mean)) %>% arrange(cluster)

```





Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.
